Inferring probabilistic stellar rotation periods using Gaussian processes

### 2017-06-21 documents/ms.tex
removing data file

\newcommand{\response}[1]{{\bf #1}}
\newcommand{\Kepler}{{\it Kepler}}
\newcommand{\kepler}{\Kepler}
\newcommand{\corot}{{\it CoRoT}}
\newcommand{\Ktwo}{{\it K2}}
\newcommand{\ktwo}{\Ktwo}
\newcommand{\TESS}{{\it TESS}}
\newcommand{\LSST}{{\it LSST}}
\newcommand{\Wfirst}{{\it Wfirst}}
\newcommand{\SDSS}{{\it SDSS}}
\newcommand{\PLATO}{{\it PLATO}}
\newcommand{\Teff}{$T_{\mathrm{eff}}$}
\newcommand{\teff}{$T_{\mathrm{eff}}$}
\newcommand{\FeH}{[Fe/H]}
\newcommand{\feh}{[Fe/H]}
\newcommand{\ie}{{\it i.e.}}
\newcommand{\eg}{{\it e.g.}}
\newcommand{\logg}{log \emph{g}}
\newcommand{\dnu}{$\Delta \nu$}
\newcommand{\numax}{$\nu_{\mathrm{max}}$}

@@

\author[R.~Angus \emph{et al.}]{
    Ruth~Angus,$^1$\thanks{Contact e-mail:
\href{mailto:ruthangus@astro.columbia.edu}{ruthangus@astro.columbia.edu}}
    Timothy Morton$^2$,
    Suzanne Aigrain$^3$,
    Daniel Foreman-Mackey$^4$,
    \newauthor
    \& Vinesh Rajpaul$^3$
    \\
    $^1$Simons Fellow, Department of Astronomy, Columbia University, NY, NY \\
    $^2$Department of Astrophysical Sciences, Princeton University,
    Princeton, NJ \\
    $^3$Subdepartment of Astrophysics, University of Oxford, UK \\
    $^4$Sagan Fellow, Department of Astronomy, University of Washington,
    Seattle, WA}

@@

\begin{abstract}
Variability in the light curves of spotted, rotating stars is often
    non-sinusoidal and quasi-periodic --- spots move on the stellar surface
    and have finite lifetimes, causing stellar flux variations to slowly shift
    in phase.
A strictly periodic sinusoid therefore cannot accurately model a rotationally
    modulated stellar light curve.
Physical models of stellar surfaces have many drawbacks preventing effective
    inference, such as highly degenerate or high-dimensional parameter spaces.
In this work, we test an appropriate {\it effective} model: a Gaussian
    Process with a quasi-periodic covariance kernel function.
This highly flexible model allows sampling of the posterior probability
    density function of the periodic parameter, marginalising over the
    other kernel hyperparameters using a Markov Chain Monte Carlo approach.
To test the effectiveness of this method, we infer rotation periods from
    \naigrain\ simulated stellar light curves, demonstrating that the Gaussian
    process method produces periods that are more accurate than both
    a sine-fitting periodogram and an autocorrelation function method.
We also demonstrate that it works well on real data, by inferring
    rotation periods for \nkoimcq\ \Kepler\ stars with previously measured
    periods.
    \response{We provide a table of rotation periods for these \nkoi\ \Kepler\ objects of
    interest and their posterior probability density function samples.}
Because this method delivers posterior probability density functions, it will
    enable hierarchical studies involving stellar rotation, particularly those
    involving population modelling, such as inferring stellar ages,
    obliquities in exoplanet systems, or characterising star-planet
    interactions.
    The code used to implement this method is available online.
% The ACF method is not applicable to unevenly spaced data,
% therefore it is inappropriate for the ~$10^7$ stellar light curves expected
% to be produced by \LSST\ unlike the GP method.
% Furthermore, the improvement is expected to be even more dramatic when applied
% to real, noisy {\it Kepler} light curves, since the GP method is well suited
% to modelling rotation signals and correlated noise simultaneously.

@@

Figure \ref{fig:mcquillan} compares the periods inferred with the GP method to
the ACF-based periods from \citet{Mcquillan2013} for the \nkoimcq\ overlapping
KOIs.
This comparison shows generally very good agreement, demonstrating that this
method works not only on simulated data, but also on real data---with the
caveat that for any particular data set, some care is needed regarding the
setting the priors and ranges for the GP hyperparameters.
% Notably, the GP method recovers periods systematically slightly larger than
% \citet{Mcquillan2013}---a likely consequence of correcting for the ACF peak
% measurement bias discussed in Section \ref{sec:acf}.
\response{The new GP rotation periods calculated for \nkoi\ KOIs are provided in an
online table accompanying this paper.
The first few lines of this table is shown below
(table \ref{tab:koi_periods}).
Samples of the rotation period posterior PDFs are available online:
\url{https://doi.org/10.5281/zenodo.804440}}

\begin{table*}
\begin{center}
    \caption{Rotation periods and \Kepler\ input catalogue (KIC) properties
    for the \Kepler\ objects of interest. The full table is available online.}
\begin{tabular}{lcccc}
    KOI number & Period (days) & KIC T$_{\mathrm{eff}}$ & KIC $\log(g)$ & KIC
    $[$Fe$/$H$]$ \\
    \hline
    1 & 25$\pm1$ & $5126^{+160}_{-138}$ & $4.61^{+0.03}_{-0.1}$ &
    $-0.1\pm0.3$ \\
    2 & 20$\pm1$ & $5850\pm50$ & $4.46\pm0.03$ & $-0.2\pm0.1$ \\
    3 & 28$\pm1$ & $6350\pm80$ & $4.02\pm0.01$ & $0.26\pm0.08$\\
    6 & 23$\pm1$ & $6225^{+114}_{-158}$ & $4.17^{+0.06}_{-0.05}$ &
    $-0.0^{+0.1}_{-0.2}$\\
    7 & 25$\pm1$ & $5543\pm79 $ & $4.08\pm0.01$ & $0.4\pm0.1$ \\
    ... & ... & ... & ... & ... \\
\end{tabular}
\end{center}
\end{table*}
\label{tab:koi_periods}

\begin{figure*}
\begin{center}
\includegraphics[width=6in, clip=true]{figures/comparison_koi_02_03.pdf}
\caption[Comparison with McQuillan results.]
{A comparison of our GP rotation period measurements to those of
\citet{Mcquillan2013}.
The data points are coloured by the range of variability measured by
    \citet{Mcquillan2013}, defined as the interval between the 5th and 95th
    percentiles of normalised flux per period bin in millimagnitudes.}
\label{fig:mcquillan}
\end{center}
\end{figure*}

>>>>

\newcommand{\response}[1]{{#1}}
\newcommand{\Kepler}{{\it Kepler}}
\newcommand{\kepler}{\Kepler}
\newcommand{\corot}{{\it CoRoT}}
\newcommand{\Ktwo}{{\it K2}}
\newcommand{\ktwo}{\Ktwo}
\newcommand{\TESS}{{\it TESS}}
\newcommand{\LSST}{{\it LSST}}
\newcommand{\Wfirst}{{\it Wfirst}}
\newcommand{\SDSS}{{\it SDSS}}
\newcommand{\PLATO}{{\it PLATO}}
\newcommand{\Teff}{$T_{\mathrm{eff}}$}
\newcommand{\teff}{$T_{\mathrm{eff}}$}
\newcommand{\FeH}{[Fe/H]}
\newcommand{\feh}{[Fe/H]}
\newcommand{\ie}{{\it i.e.}}
\newcommand{\eg}{{\it e.g.}}
\newcommand{\logg}{log \emph{g}}
\newcommand{\dnu}{$\Delta \nu$}
\newcommand{\numax}{$\nu_{\mathrm{max}}$}

@@

\author[R.~Angus \emph{et al.}]{
    Ruth~Angus,$^1$\thanks{Contact e-mail:
\href{mailto:ruthangus@astro.columbia.edu}{ruthangus@astro.columbia.edu}}
    Timothy Morton$^2$,
    Suzanne Aigrain$^3$,
    Daniel Foreman-Mackey$^4$
    \newauthor
    \& Vinesh Rajpaul$^3$
    \\
    $^1$Simons Fellow, Department of Astronomy, Columbia University, NY, NY \\
    $^2$Department of Astrophysical Sciences, Princeton University,
    Princeton, NJ \\
    $^3$Subdepartment of Astrophysics, University of Oxford, UK \\
    $^4$Sagan Fellow, Department of Astronomy, University of Washington,
    Seattle, WA}

@@

\begin{abstract}
Variability in the light curves of spotted, rotating stars is often
    non-sinusoidal and quasi-periodic --- spots move on the stellar surface
    and have finite lifetimes, causing stellar flux variations to slowly shift
    in phase.
A strictly periodic sinusoid therefore cannot accurately model a rotationally
    modulated stellar light curve.
Physical models of stellar surfaces have many drawbacks preventing effective
    inference, such as highly degenerate or high-dimensional parameter spaces.
In this work, we test an appropriate {\it effective} model: a Gaussian
    Process with a quasi-periodic covariance kernel function.
This highly flexible model allows sampling of the posterior probability
    density function of the periodic parameter, marginalising over the
    other kernel hyperparameters using a Markov Chain Monte Carlo approach.
To test the effectiveness of this method, we infer rotation periods from
    \naigrain\ simulated stellar light curves, demonstrating that the Gaussian
    process method produces periods that are more accurate than both
    a sine-fitting periodogram and an autocorrelation function method.
We also demonstrate that it works well on real data, by inferring
    rotation periods for \nkoimcq\ \Kepler\ stars with previously measured
    periods.
    % \response{We provide a table of rotation periods for these \nkoi\ \Kepler\ objects of
    % interest and their posterior probability density function samples.}
Because this method delivers posterior probability density functions, it will
    enable hierarchical studies involving stellar rotation, particularly those
    involving population modelling, such as inferring stellar ages,
    obliquities in exoplanet systems, or characterising star-planet
    interactions.
    The code used to implement this method is available online.
% The ACF method is not applicable to unevenly spaced data,
% therefore it is inappropriate for the ~$10^7$ stellar light curves expected
% to be produced by \LSST\ unlike the GP method.
% Furthermore, the improvement is expected to be even more dramatic when applied
% to real, noisy {\it Kepler} light curves, since the GP method is well suited
% to modelling rotation signals and correlated noise simultaneously.

@@

Figure \ref{fig:mcquillan} compares the periods inferred with the GP method to
the ACF-based periods from \citet{Mcquillan2013} for the \nkoimcq\ overlapping
KOIs.
This comparison shows generally very good agreement, demonstrating that this
method works not only on simulated data, but also on real data---with the
caveat that for any particular data set, some care is needed regarding the
setting the priors and ranges for the GP hyperparameters.
% Notably, the GP method recovers periods systematically slightly larger than
% \citet{Mcquillan2013}---a likely consequence of correcting for the ACF peak
% measurement bias discussed in Section \ref{sec:acf}.

% \response{The new GP rotation periods calculated for \nkoi\ KOIs are provided in an
% online table accompanying this paper.
% The first few lines of this table is shown below
% (table \ref{tab:koi_periods}).
% Samples of the rotation period posterior PDFs are available online:
% \url{https://doi.org/10.5281/zenodo.804440}}

% \begin{table*}
% \begin{center}
%     \caption{Rotation periods and \Kepler\ input catalogue (KIC) properties
%     for the \Kepler\ objects of interest. The full table is available online.}
% \begin{tabular}{lcccc}
%     KOI number & Period (days) & KIC T$_{\mathrm{eff}}$ & KIC $\log(g)$ & KIC
%     $[$Fe$/$H$]$ \\
%     \hline
%     1 & 25$\pm1$ & $5126^{+160}_{-138}$ & $4.61^{+0.03}_{-0.1}$ &
%     $-0.1\pm0.3$ \\
%     2 & 20$\pm1$ & $5850\pm50$ & $4.46\pm0.03$ & $-0.2\pm0.1$ \\
%     3 & 28$\pm1$ & $6350\pm80$ & $4.02\pm0.01$ & $0.26\pm0.08$\\
%     6 & 23$\pm1$ & $6225^{+114}_{-158}$ & $4.17^{+0.06}_{-0.05}$ &
%     $-0.0^{+0.1}_{-0.2}$\\
%     7 & 25$\pm1$ & $5543\pm79 $ & $4.08\pm0.01$ & $0.4\pm0.1$ \\
%     ... & ... & ... & ... & ... \\
% \end{tabular}
% \end{center}
% \end{table*}
% \label{tab:koi_periods}

\begin{figure*}
\begin{center}
\includegraphics[width=6in, clip=true]{figures/comparison_koi_01_23.pdf}
\caption[Comparison with McQuillan results.]
{A comparison of our GP rotation period measurements to those of
\citet{Mcquillan2013}.
The data points are coloured by the range of variability measured by
    \citet{Mcquillan2013}, defined as the interval between the 5th and 95th
    percentiles of normalised flux per period bin in millimagnitudes.}
\label{fig:mcquillan}
\end{center}
\end{figure*}
